### 2025-03-12 paper/sections/method.qmd
Review counterfactual training up to 3.4

Counterfactual training combines ideas from adversarial training, energy-based modelling and counterfactuals explanations with the explicit objective of aligning representations with plausible explanations that comply with user requirements. In the context of CEs, plausibility has broadly been defined as the degree to which counterfactuals comply with the underlying data generating process [@poyiadzi2020face;@guidotti2022counterfactual;@altmeyer2024faithful]. Plausibility is a necessary but insufficient condition for using CEs to provide algorithmic recourse (AR) to individuals affected by opaque models in practice. This is because for recourse recommendations to be **actionable**, they need to not only result in plausible counterfactuals but also be attainable. A plausible CE for a rejected 20-year-old loan applicant, for example, might reveal that their application would have been accepted, if only they were 20 years older. Ignoring all other features, this would comply with the definition of plausibility if 40-year-old individuals were in fact more credit-worthy on average than young adults. But of course this CE does not qualify for providing actionable recourse to the applicant since *age* is not a (directly) mutable feature. For our intents and purposes, counterfactual training aims to improve model explainability by aligning models with counterfactuals that meet both desiderata, plausibility and actionability. Formally, we define explainability as follows:

@@

We say that $\mathbf{M}_\theta$ is **explainable** to the extent that faithfully generated counterfactuals are plausible (i.e. consistent with the data) and actionable. Formally, we define these properties as follows:

@@

The definitions of faithfulness and plausibility in @def-explainability are the same as in @altmeyer2024faithful, with adapted notation. Actionability constraints in @def-explainability vary and depend on the context in which $\mathbf{M}_\theta$ is deployed. In this work, we focus on domain and mutability constraints for individual features $x_d$ for $d=1,...,D$. We limit ourselves to classification tasks for reasons discussed in @sec-discussion.

@@

Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote a counterfactual explanation generated through gradient descent over $T$ iterations as initially proposed by @wachter2017counterfactual. For our purposes, we let $T$ vary and consider the counterfactual search as converged as soon as the predicted probability for the target class has reached a pre-determined threshold, $\tau$: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.^[For detailed background information on gradient-based counterfactual search and convergence see @sec-app-ce.]

@@

$$
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y})
$$ {#eq-obj}

where $\text{yloss}(\cdot)$ is any conventional classification loss that induces discriminative performance (e.g., cross-entropy). The two additional components in @eq-obj are explained in more detail below. For now, they can be sufficiently described as inducing explainability directly and indirectly by penalizing: (1) the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $x$ and, (2) the adversarial loss, $\text{advloss}(.)$, with respect to nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$. The tradeoff between the different components can be governed by adjusting the strengths of the penalties $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$.

### Directly Inducing Explainability through Contrastive Divergence

@grathwohl2020your observe that any classifier can be re-interpreted as a joint energy-based model (JEM) that learns to discriminate output classes conditional on the observed (training) samples from $p(\mathbf{x})$ and the generated samples from $p_\theta(\mathbf{x})$. They show that JEMs can be trained to perform well at both tasks by directly maximizing the joint log-likelihood factorized as $\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})$. The first factor can be optimized using conventional cross-entropy as in @eq-obj. Then, to optimize $\log p_\theta(\mathbf{x})$ @grathwohl2020your minimize the contrastive divergence between these observed samples from $p(\mathbf{x})$ and generated samples from $p_\theta(\mathbf{x})$.

A key empirical finding in @altmeyer2024faithful was that JEMs tend to do well with respect to the plausibility objective in @def-explainability. If we consider samples drawn from $p_\theta(\mathbf{x})$ as counterfactuals, this is an expected finding, because the JEM objective effectively minimizes the divergence between the conditional posterior and $p(\mathbf{x}|\mathbf{y}^{+})$. To generate samples, @grathwohl2020your rely on Stochastic Gradient Langevin Dynamics (SGLD) using an uninformative prior for initialization. This is where we depart from their methodology: instead of SGLD, we propose to use counterfactual explainers to generate counterfactuals of observed training samples. Specifically, we have:

@@

where $\mathcal{E}_\theta(\cdot)$ denotes the energy function. In particular, we set $\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x^+})[y^+]$ where $y^+$ denotes the index of the randomly drawn target class, $y^+ \sim p(y)$, and $\mathbf{x^+}$ denotes an observed data point sampled from target domain: $\mathbf{X}^+=\{\mathbf{x}:y=y^+\}$. Conditional on the target class $y^+$, $\mathbf{x}_T^\prime$ denotes a mature counterfactual for a randomly sampled factual from a non-target class generated through a gradient-based counterfactual generator for at most $T$ iterations. We define mature counterfactuals as those that have either exhausted $T$ or reached convergence in terms of the pre-determined decision threshold earlier.

Intuitively, the gradient of @eq-div decreases the energy of observed training samples (positive samples) while at same time increasing the energy of counterfactuals (negative samples) [@du2019implicit]. As the generated counterfactuals get more plausible (@def-explainability) over the course of training, these two opposing effects gradually balance each out [@lippe2024uvadlc].

The departure from SGLD allows us to tap into the vast repertoire of explainers that have been proposed in the literature to meet different desiderata. Typically, these methods facilitate the imposition of domain and mutability constraints, for example. In principle, any existing approach for generating counterfactual explanations is viable, so long as it does not violate the faithfulness condition. Like JEMs [@murphy2022probabilistic], counterfactual training can be considered as a form of contrastive representation learning.

### Indirectly Inducing Explainability through Adversarial Robustness

Based on our analysis in @sec-lit, counterfactuals $\mathbf{x}^\prime$ can be repurposed as additional training samples [@luu2023counterfactual;@balashankar2023improving] or adversarial examples [@freiesleben2022intriguing;@pawelczyk2022exploring]. This leaves some flexibility with respect to the exact choice for $\text{advloss}(\cdot)$ in @eq-obj. An intuitive functional form to use, though likely not the only reasonable choice, is inspired by adversarial training:

@@

Under this choice, we consider nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$ as adversarial examples as long as the magnitude of the perturbation to any individual feature is at most $\varepsilon$. This is closely aligned with @szegedy2013intriguing, who define an adversarial attack as an "imperceptible non-random perturbation". Thus, we choose to work with a different distinction between CE and AE than @freiesleben2022intriguing, who considers misclassification as the key distinguishing feature of AE. One of the key observations in this work is that we can leverage counterfactual explanations during training and get adversarial examples, essentially for free.

>>>>

Counterfactual training (CT) combines ideas from adversarial training, energy-based modelling and counterfactuals explanations with the explicit goal of aligning representations with plausible explanations that comply with user requirements. In the context of CEs, plausibility has broadly been defined as the degree to which counterfactuals comply with the underlying data-generating process [@poyiadzi2020face;@guidotti2022counterfactual;@altmeyer2024faithful]. Plausibility is a necessary but insufficient condition for using CEs to provide algorithmic recourse (AR) to individuals (negatively) affected by opaque models. For AR recommendations to be actionable, they need to not only result in plausible counterfactuals but also be attainable. A plausible CE for a rejected 20-year-old loan applicant, for example, might reveal that their application would have been accepted, if only they were 20 years older. Ignoring all other features, this would comply with the definition of plausibility if 40-year-old individuals were in fact more credit-worthy on average than young adults. But of course this CE does not qualify for providing actionable recourse to the applicant since *age* is not a (directly) mutable feature. CT aims to improve model explainability by aligning models with counterfactuals that meet both desiderata: plausibility and actionability. Formally, we define explainability as follows:

@@

We say that $\mathbf{M}_\theta$ is **explainable** to the extent that faithfully generated counterfactuals are plausible (i.e., consistent with the data) and actionable. Formally, we define these properties as follows:

@@

The characterization of faithfulness and plausibility in @def-explainability is the same as in @altmeyer2024faithful, with adapted notation. Actionability constraints in @def-explainability vary and depend on the context in which $\mathbf{M}_\theta$ is deployed. In this work, we focus on domain and mutability constraints for individual features $x_d$ for $d=1,...,D$. We limit ourselves to classification tasks for reasons discussed in @sec-discussion.

@@

Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote a counterfactual explanation generated through gradient descent over $T$ iterations as initially proposed by @wachter2017counterfactual. For our purposes, we let $T$ vary and consider the counterfactual search to be converged as soon as the predicted probability for the target class has reached a pre-determined threshold $\tau$: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.^[For detailed background information on gradient-based counterfactual search and convergence see @sec-app-ce.]

@@

$$
\begin{split}
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) &+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) \\ &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta)
\end{split}
$$ {#eq-obj}

where $\text{yloss}(\cdot)$ is a classification loss that induces discriminative performance (e.g., cross-entropy). The second and third terms in @eq-obj are explained in detail below. For now, they can be sufficiently described as inducing explainability directly and indirectly by penalizing: (1) the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $x$ and, (2) the adversarial loss, $\text{advloss}(.)$, with respect to nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$. Finally, $\text{ridge}(\cdot)$ denotes a Ridge penalty ($\ell_2$-norm) that regularizes the magnitude of the energy terms involved in $\text{div}(\cdot)$ [@du2019implicit]. The trade-off between the components can be governed by adjusting the strengths of the penalties $\lambda_{\text{div}}$, $\lambda_{\text{adv}}$ and $\lambda_{\text{reg}}$.

## Directly Inducing Explainability with Contrastive Divergence

@grathwohl2020your observe that any classifier can be re-interpreted as a joint energy-based model (JEM) that learns to discriminate output classes conditional on the observed (training) samples from $p(\mathbf{x})$ and the generated samples from $p_\theta(\mathbf{x})$. The authors show that JEMs can be trained to perform well at both tasks by directly maximizing the joint log-likelihood factorized as $\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})$. The first term can be optimized using conventional cross-entropy as in @eq-obj. Then, to optimize $\log p_\theta(\mathbf{x})$ @grathwohl2020your minimize the contrastive divergence between these observed samples from $p(\mathbf{x})$ and generated samples from $p_\theta(\mathbf{x})$.

A key empirical finding in @altmeyer2024faithful was that JEMs tend to do well with respect to the plausibility objective in @def-explainability. This follows directly if we consider samples drawn from $p_\theta(\mathbf{x})$ as counterfactuals because the JEM objective effectively minimizes the divergence between the conditional posterior and $p(\mathbf{x}|\mathbf{y}^{+})$. To generate samples, @grathwohl2020your rely on Stochastic Gradient Langevin Dynamics (SGLD) using an uninformative prior for initialization but we depart from their methodology. Instead of SGLD, we propose to use counterfactual explainers to generate counterfactuals of observed training samples. Specifically, we have:

@@

where $\mathcal{E}_\theta(\cdot)$ denotes the energy function. We set $\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x^+})[y^+]$ where $y^+$ denotes the index of the randomly drawn target class, $y^+ \sim p(y)$, and $\mathbf{x^+}$ denotes an observed sample from target domain: $\mathbf{X}^+=\{\mathbf{x}:y=y^+\}$. Conditional on the target class $y^+$, $\mathbf{x}_T^\prime$ denotes a mature counterfactual for a randomly sampled factual from a non-target class generated with a gradient-based CE generator for up to $T$ iterations. Mature counterfactuals are ones that have either reached convergence wrt. the decision threshold $\tau$ or exhausted $T$.

Intuitively, the gradient of @eq-div decreases the energy of observed training samples (positive samples) while increasing the energy of counterfactuals (negative samples) [@du2019implicit]. As the counterfactuals get more plausible (@def-explainability) during training, these opposing effects gradually balance each other out [@lippe2024uvadlc].

The departure from SGLD allows us to tap into the vast repertoire of explainers that have been proposed in the literature to meet different desiderata. For example, many methods facilitate the imposition of domain and mutability constraints. In principle, any existing approach for generating counterfactual explanations is viable, so long as it does not violate the faithfulness condition. Like JEMs [@murphy2022probabilistic], CT can be considered a form of contrastive representation learning.

## Indirectly Inducing Explainability with Adversarial Robustness

Based on our analysis in @sec-lit, counterfactuals $\mathbf{x}^\prime$ can be repurposed as additional training samples [@luu2023counterfactual;@balashankar2023improving] or AEs [@freiesleben2022intriguing;@pawelczyk2022exploring]. This leaves some flexibility with respect to the choice for $\text{advloss}(\cdot)$ in @eq-obj. An intuitive functional form, but likely not the only sensible choice, is inspired by adversarial training:

@@

Under this choice, we consider nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$ as AEs as long as the magnitude of the perturbation to any single feature is at most $\varepsilon$. This is closely aligned with @szegedy2013intriguing who define an adversarial attack as an "imperceptible non-random perturbation". Thus, we choose to work with a different distinction between CE and AE than @freiesleben2022intriguing who consider misclassification as the key distinguishing feature of AE. One of the key observations in this work is that we can leverage CEs during training and get adversarial examples essentially for free.
